---
title: "Reconstructing the Composite Metabolic Desirability Index after a triglyceride data correction"
subtitle: "Diagnostics for the index construction prior to refitting the prediction model"
date: now
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: environment
#| eval: true
load("/Users/macbookpro/Desktop/Rprojects/codiet3/.RData")                                           
```
```{r}
#| label: setup
#| eval: true

library(tidyverse)
library(patchwork)
# library(readxl)
library(Hmisc)
library(desirability2)
library(knitr)
```

# Purpose and scope

Triglyceride values were found to be inaccurate for a substantial share of
observations in the biochemical panel. Because triglycerides are one of the four
components from which the Composite Metabolic Desirability Index (CMDI) is built,
the correction propagates into the outcome of the clinical prediction model
rather than into a single predictor.

The prediction model is fitted with `bootPriorProc`, which takes approximately
6.2 hours to run. Repeating that run under several candidate constructions would
be costly, so the purpose of this document is to settle the index construction
first, using diagnostics that are less time consuming.

Three constructions are compared:

| Label | Triglycerides | Desirability bounds |
|---|---|---|
| **v1** | original | data-driven (`use_data = TRUE`) |
| **v2** | corrected | data-driven (`use_data = TRUE`) |
| **v3** | corrected | fixed clinical bounds (`use_data = FALSE`) |

The three pairwise comparisons separate the two sources of change: v1 → v2
isolates the data correction, v2 → v3 isolates the parameterisation, and v1 → v3
gives the total that downstream reporting would have to absorb.

The prediction model is an ordinal regression one, which is
rank-based and invariant to monotone transformations of the outcome. Changes in
the *scale* or *spread* of the CMDI therefore have no effect on the fitted model;
only changes in the *ordering* of participants do. Rank correlation is used
throughout as the primary measure of impact for this reason.

# Data preparation and the three index constructions

## Corrected triglycerides and imputation

```{r}
#| label: load-tg

tg_corrected <- readxl::read_xlsx(
  "/Users/macbookpro/Desktop/Rprojects/codiet2/dat/latest2_triglycerides_change/WP2 CODIET BIOCHEMICAL_HEMATOLOGICAL_INDICES FINAL JULY CHECK  31.7.26.xlsx",
  sheet = "biochem_310726")

tg_corrected <- tg_corrected |>
  separate(ID, c("x", "ID", "t"), sep = "_") |>
  select(-x, -t) |>
  select(-Site_Collection)

biochem_TGcorr <- biochem |>
  select(-TRIG) |>
  inner_join(tg_corrected, by = c("ID", "Timepoint")) |>
  relocate(TRIG, .before = HDL)
```

Missing values are imputed with the same random-forest procedure used for the
original analysis (`miceRanger`), and the multiply-imputed datasets are collapsed
by taking the mean for numeric variables and the mode for character variables.

```{r}
#| label: impute

# mR2_biochem_TGcorr is produced by running for_jobs.R

fMode      <- function(x) names(which.max(table(x)))
fsummarise <- function(x) if (is.character(x)) fMode(x) else round(mean(x), 2)

imp_dataList <- miceRanger::completeData(mR2_biochem_TGcorr)
imp_dataList <- map(imp_dataList, \(x) x |> rownames_to_column(var = "id")) |>
  list_rbind()

d_imp <- imp_dataList[, lapply(.SD, fsummarise), by = id]
biochem_TGcorr_imp <- d_imp |> select(-id) |> as_tibble()
rm(d_imp, imp_dataList, fsummarise, fMode)

bp_rf_input <- human_dat_imp |>
  filter(Timepoint %in% c(1, 3)) |>
  unite("ID_Timepoint", ID, Timepoint) |>
  select(ID_Timepoint, bmi) |>
  inner_join(
    biochem_TGcorr_imp |> unite("ID_Timepoint", ID, Timepoint) |>
      select(ID_Timepoint, TRIG, HDL, GLU),
    by = "ID_Timepoint") |>
  relocate(bmi, .before = TRIG) |>
  separate(ID_Timepoint, into = c("ID", "Timepoint"))
```

## Scale of the triglyceride change

The correction is not marginal. Median triglycerides rise by half and the upper
tail extends considerably further.

| Statistic | Original | Corrected |
|---|---:|---:|
| n | 299 | 296 |
| Mean | 84.6 | 123.7 |
| Median | 70.0 | 105.0 |
| Gini mean difference | 57.5 | 79.7 |
| 25th percentile | 46.8 | 70.9 |
| 75th percentile | 107.0 | 141.7 |
| 95th percentile | 180.2 | 235.0 |
| Maximum | 590.1 | 1351.6 |

The difference in row counts reflects three records with missing BMI, which the
desirability functions exclude in either case; the analysis set is unaffected.

## v2: corrected triglycerides, data-driven bounds

This is the construction as originally specified, with the triglyceride `scale`
re-tuned so that the IDF threshold of 150 mg/dL continues to map to a desirability
of 0.6 under the new data range.

```{r}
#| label: build-v2

bp_rf_output_reducedIndex_TGcorr <- bp_rf_input |> mutate(
  d_BMI  = d_max(bmi,  scale = 0.5015, use_data = TRUE),   # target 30 >
  d_TRIG = d_max(TRIG, scale = 0.2095, use_data = TRUE),   # target 150 >= (re-tuned)
  d_HDL  = d_min(HDL,  scale = 2.08,   use_data = TRUE),   # target 45 <
  d_GLU  = d_max(GLU,  scale = 0.8135, use_data = TRUE),   # target 100 >=
  ms_score = round(100 * d_overall(across(starts_with("d_"))), digits = 4)
)
```

## v3: corrected triglycerides, fixed clinical bounds

The bounds are stated explicitly and the `scale` for each component is solved so
that the clinical anchor is preserved exactly. Solving rather than hard-coding
means a change to any bound cannot silently break the anchoring.

```{r}
#| label: bounds

BOUNDS <- tibble::tribble(
  ~comp,  ~kind,  ~low,  ~high, ~anchor, ~target,
  "BMI",  "max",  14,    70,     30,     0.5,
  "TRIG", "max",  15,    1000,   150,    0.5,
  "HDL",  "min",  10,    130,    45,     0.5,
  "GLU",  "max",  40,    500,    100,    0.5
)

scale_for <- function(anchor, low, high, target, kind) {
  frac <- if (kind == "max") (anchor - low) / (high - low)
          else               (high - anchor) / (high - low)
  stopifnot(frac > 0, frac < 1)
  log(target) / log(frac)
}

BOUNDS <- BOUNDS |>
  rowwise() |>
  mutate(scale = scale_for(anchor, low, high, target, kind)) |>
  ungroup()

getb <- function(cp, f) BOUNDS[[f]][BOUNDS$comp == cp]
BOUNDS
```

| Component | Kind | low | high | anchor | target | scale |
|---|---|---:|---:|---:|---:|---:|
| BMI | `d_max` | 14 | 70 | 30 | 0.5 | 0.553 |
| TRIG | `d_max` | 15 | 1000 | 150 | 0.5 | 0.349 |
| HDL | `d_min` | 10 | 130 | 45 | 0.5 | 2.010 |
| GLU | `d_max` | 40 | 500 | 100 | 0.5 | 0.340 |

A small floor is applied to each component before aggregation. Under a geometric
mean a single component at zero forces the whole index to zero, so the floor
prevents one extreme value from erasing the information in the other three. With
the bounds above no observation reaches the floor, so it acts as insurance rather
than doing real work.

```{r}
#| label: build-v3

EPS <- 0.01

bp_rf_output_reducedIndex_fixed <- bp_rf_input |>
  mutate(
    d_BMI  = d_max(bmi,  low = getb("BMI","low"),  high = getb("BMI","high"),
                   scale = getb("BMI","scale"),  use_data = FALSE),
    d_TRIG = d_max(TRIG, low = getb("TRIG","low"), high = getb("TRIG","high"),
                   scale = getb("TRIG","scale"), use_data = FALSE),
    d_HDL  = d_min(HDL,  low = getb("HDL","low"),  high = getb("HDL","high"),
                   scale = getb("HDL","scale"),  use_data = FALSE),
    d_GLU  = d_max(GLU,  low = getb("GLU","low"),  high = getb("GLU","high"),
                   scale = getb("GLU","scale"),  use_data = FALSE)
  ) |>
  mutate(n_zero_raw = rowSums(across(starts_with("d_"), ~ .x <= 0), na.rm = TRUE)) |>
  mutate(across(starts_with("d_"), ~ pmax(.x, EPS))) |>
  mutate(ms_score = round(100 * d_overall(across(starts_with("d_"))), digits = 4))
```

# Why the bounds were changed

The move from data-driven to fixed bounds was not part of the original plan. It
was prompted by two observations that emerged while diagnosing the correction.

## Scores that depend on other participants

With `use_data = TRUE` the desirability bounds are taken from the observed range,
so the participant holding the sample minimum receives a desirability of exactly
zero on that component. Because the index is a geometric mean, that zero
propagates to the whole index.

The clearest example is ID 132 at timepoint 3. Their triglyceride value is 35
mg/dL in both the original and corrected datasets, it did not change. Their CMDI
nonetheless fell from 43.5 to 0, because the sample minimum moved from 15.47 up
to 35, making them the new floor-holder.

| ID | Timepoint | TRIG (v1) | TRIG (v2) | CMDI v1 | CMDI v2 |
|---|---|---:|---:|---:|---:|
| 132 | 3 | 35.0 | 35.0 | 43.5 | 0.0 |
| 207 | 1 | 15.5 | 35.4 | 0.0 | 19.6 |

An index in which one participant's score changes because a different
participant's data changed is difficult to defend as an individual-level
quantity, and it is incompatible with applying the index to an external cohort.

## A clinically implausible score

ID 132 at timepoint 3 has BMI 31.1, HDL 50, glucose 96 and triglycerides 35. A
CMDI of zero places this participant at the most metabolically desirable end of
the scale, which is not a reasonable description of someone in the obese BMI
range. The zero is an artefact of the floor mechanism rather than a summary of
their metabolic state.

## Selecting the bounds

Three criteria were used, in order:

1. Clinical defensibility: Bounds anchored to published thresholds and chosen
   a priori rather than from the observed data.
2. Coverage: Wide enough that saturation is rare, verified empirically. This
   matters because a saturated value carries no information: every observation
   beyond a bound receives an identical desirability.
3. Portability: Fixed, so that the index can be computed identically in an
   external cohort.

The bounds were widened iteratively. An initial set (BMI 16–55, TG 20–500, HDL
20–110, glucose 55–250) produced 23 observations at the floor, driven by HDL ≥ 80
and glucose ≤ 70 — both common values rather than clinical extremes. The final
set places each bound beyond the plausible clinical range: severe underweight
through super-obesity for BMI, values below any plausible fasting measurement
through the threshold for pancreatitis risk for triglycerides, severe
hypoalphalipoproteinaemia through exceptional hyperalphalipoproteinaemia for HDL,
and clinical hypoglycaemia through marked hyperglycaemia for glucose.

## Saturation audit

```{r}
#| label: saturation

bp_rf_output_reducedIndex_fixed |>
  summarise(
    BMI_lo  = sum(bmi  <= getb("BMI","low")),   BMI_hi  = sum(bmi  >= getb("BMI","high")),
    TRIG_lo = sum(TRIG <= getb("TRIG","low")),  TRIG_hi = sum(TRIG >= getb("TRIG","high")),
    HDL_lo  = sum(HDL  <= getb("HDL","low")),   HDL_hi  = sum(HDL  >= getb("HDL","high")),
    GLU_lo  = sum(GLU  <= getb("GLU","low")),   GLU_hi  = sum(GLU  >= getb("GLU","high"))
  ) |>
  kable(caption = "Observations reaching each bound")
```

| BMI_lo | BMI_hi | TRIG_lo | TRIG_hi | HDL_lo | HDL_hi | GLU_lo | GLU_hi |
|---:|---:|---:|---:|---:|---:|---:|---:|
| 0 | 0 | 0 | 1 | 0 | 0 | 0 | 0 |

No observation reaches a lower bound, so the zero-collapse mechanism cannot
operate. One observation sit at or above the triglyceride ceiling of 1000
mg/dL. This is acceptable: values in that range fall into a clinically homogeneous
category in which management does not differentiate further, so collapsing them
is a reasonable simplification rather than a loss.

# The anchor value

Each component's `target` states the desirability assigned to a participant
sitting exactly at the IDF threshold. The original specification used 0.5 where
the IDF states a criterion with a strict inequality and 0.6 where it uses a
non-strict inequality, on the reasoning that a participant at the threshold has
already met the criterion in the latter case.

The direction of that adjustment follows from the definition, but the magnitude
does not. The distinction between `>150` and `≥150` concerns a single point on a
continuous scale, and encoding it as a 0.1 shift in desirability, which
propagates to a 36% change in the scale parameter, gives it more weight than it
can carry. Two of the four anchors are also not clean IDF thresholds: the HDL
anchor of 45 mg/dL is a sex-neutral compromise between the 40 mg/dL and 50 mg/dL
criteria, and BMI > 30 is the IDF's surrogate provision for when waist
circumference is unavailable rather than the primary criterion.

A single anchor is used for all four components instead. The justification is
that the IDF assigns equal status to each criterion (each contributes one
qualifying condition, none counts for more than another) so mapping all four
thresholds to the same desirability preserves that equivalence.

## The specific value does not matter

With a uniform target $c$ across components, the scale for component $i$ is
$s_i = \log c / \log f_{a,i}$ where $f_{a,i}$ is the anchor's fractional position
within the bounds. Then

$$\log d_i = s_i \log f_i = \log c \cdot \frac{\log f_i}{\log f_{a,i}}$$

and the log of the aggregated index is

$$\log \text{CMDI} = \frac{1}{4}\sum_i \log d_i = \log c \cdot K, \qquad
K = \frac{1}{4}\sum_i \frac{\log f_i}{\log f_{a,i}}$$

$K$ does not involve $c$, so $\text{CMDI} = c^{K}$ is a strictly monotone function
of the same underlying quantity for any $c \in (0,1)$. Participant ordering is
therefore identical for every uniform anchor, and because all component
influences scale by $|\log c|$ together, the relative weighting is unchanged as
well.

```{r}
#| label: anchor-invariance
#| eval: true

cmdi_with_target <- function(dat, BOUNDS, c_target) {
  B <- BOUNDS |> mutate(target = c_target) |>
    rowwise() |> mutate(scale = scale_for(anchor, low, high, target, kind)) |> ungroup()
  gb <- function(cp, f) B[[f]][B$comp == cp]
  dat |>
    transmute(
      d_BMI  = d_max(bmi,  low = gb("BMI","low"),  high = gb("BMI","high"),
                     scale = gb("BMI","scale"),  use_data = FALSE),
      d_TRIG = d_max(TRIG, low = gb("TRIG","low"), high = gb("TRIG","high"),
                     scale = gb("TRIG","scale"), use_data = FALSE),
      d_HDL  = d_min(HDL,  low = gb("HDL","low"),  high = gb("HDL","high"),
                     scale = gb("HDL","scale"),  use_data = FALSE),
      d_GLU  = d_max(GLU,  low = gb("GLU","low"),  high = gb("GLU","high"),
                     scale = gb("GLU","scale"),  use_data = FALSE)
    ) |>
    mutate(across(everything(), ~ pmax(.x, EPS))) |>
    transmute(ms = 100 * d_overall(across(starts_with("d_")))) |>
    pull(ms)
}

anchor_check <- tibble(c_target = c(0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8)) |>
  mutate(
    idx      = map(c_target, ~ cmdi_with_target(bp_rf_input, BOUNDS, .x)),
    median   = map_dbl(idx, median),
    min      = map_dbl(idx, min),
    max      = map_dbl(idx, max),
    spearman = map_dbl(idx, ~ cor(.x, cmdi_with_target(bp_rf_input, BOUNDS, 0.5),
                                  method = "spearman"))
  ) |>
  select(-idx)

anchor_check |> kable(digits = c(1, 2, 2, 2, 6))
```

The rank correlation against the reference is exactly 1 for every value tested,
confirming that the choice is a labelling decision. A value of 0.5 is used as the
midpoint of the scale, and the `Methods` section of the paper can state plainly that 
the specific value is inconsequential.

# Assembling the comparison

```{r}
#| label: assemble
#| eval: true

key_int <- function(d) d |> mutate(across(c(ID, Timepoint), as.integer))

pull_v <- function(d, tag)
  key_int(d) |>
  select(ID, Timepoint, starts_with("d_"), ms_score) |>
  rename_with(~ paste0(.x, "_", tag), c(starts_with("d_"), "ms_score"))

cmp <- temp_dat |>
  key_int() |>
  select(ID, Timepoint) |>
  left_join(pull_v(bp_rf_output_reducedIndex,        "v1"), by = c("ID","Timepoint")) |>
  left_join(pull_v(bp_rf_output_reducedIndex_TGcorr, "v2"), by = c("ID","Timepoint")) |>
  left_join(pull_v(bp_rf_output_reducedIndex_fixed,  "v3"), by = c("ID","Timepoint")) |>
  filter(!is.na(ms_score_v1), !is.na(ms_score_v2), !is.na(ms_score_v3))

n_tot <- nrow(cmp)

CUT <- c(v1 = 44.29,
         v2 = median(cmp$ms_score_v2),
         v3 = median(cmp$ms_score_v3))

cmp <- cmp |>
  mutate(
    str_v1 = ifelse(ms_score_v1 >= CUT["v1"], "high", "low"),
    str_v2 = ifelse(ms_score_v2 >= CUT["v2"], "high", "low"),
    str_v3 = ifelse(ms_score_v3 >= CUT["v3"], "high", "low")
  )
```

The high-risk cut-point is the median of each index. Because the cut is defined
this way, each version splits the cohort in half by construction, so changes in
stratum membership reflect rank movement rather than a shift in the overall
distribution.

# Distribution of the three indices

```{r}
#| label: distribution-table
#| eval: true

gmd <- function(x) Hmisc::GiniMd(x[!is.na(x)])

summ <- tibble(
  version = c("v1 original", "v2 TG-corrected", "v3 TG-corr + fixed bounds"),
  cut     = as.numeric(CUT),
  median  = c(median(cmp$ms_score_v1), median(cmp$ms_score_v2), median(cmp$ms_score_v3)),
  mean    = c(mean(cmp$ms_score_v1),   mean(cmp$ms_score_v2),   mean(cmp$ms_score_v3)),
  gmd     = c(gmd(cmp$ms_score_v1),    gmd(cmp$ms_score_v2),    gmd(cmp$ms_score_v3)),
  min     = c(min(cmp$ms_score_v1),    min(cmp$ms_score_v2),    min(cmp$ms_score_v3)),
  max     = c(max(cmp$ms_score_v1),    max(cmp$ms_score_v2),    max(cmp$ms_score_v3)),
  n_zero  = c(sum(cmp$ms_score_v1 == 0), sum(cmp$ms_score_v2 == 0),
              sum(cmp$ms_score_v3 == 0))
)

summ |> kable(digits = 2)
```

Descriptive statistics for the final construction over the full set of 296
scored observations:

| Statistic | Value |
|---|---:|
| n | 296 |
| Distinct values | 296 |
| Mean | 45.68 |
| Median | 45.98 |
| Gini mean difference | 6.99 |
| 5th percentile | 34.48 |
| 25th percentile | 41.54 |
| 75th percentile | 50.25 |
| 95th percentile | 54.34 |
| Minimum | 27.99 |
| Maximum | 66.97 |

Two features are worth commenting on.

The index is more compressed than under data-driven bounds, with 90% of
participants between roughly 34 and 54. This is expected: the bounds now
represent clinically extreme values, and a score near 90 would require a
participant to sit near the pathological extreme of all four components at once.
Under a geometric mean, four moderate desirabilities produce a moderate
composite. Since `ordinal regression` is invariant to monotone transformation, the compression
has no effect on the model.

The maximum is also lower than under data-driven bounds. This is arguably more
honest: `use_data = TRUE` places the worst participant near the top of the scale
by construction, which gives a misleading impression that the cohort spans the
full range of metabolic risk. Fixed anchors make it visible that it does not,
which is consistent with the limitation already noted in the manuscript.

```{r}
#| label: fig-distributions
#| fig-cap: "CMDI distributions under the three constructions, with the published cut-point and the re-derived cut-point marked."
#| fig-width: 6
#| fig-height: 7
#| eval: true

theme_set(ggstatsplot::theme_ggstatsplot()) 

long3 <- cmp |>
  select(ms_score_v1, ms_score_v2, ms_score_v3) |>
  pivot_longer(everything(), names_to = "version", values_to = "cmdi") |>
  mutate(version = recode(version,
                          ms_score_v1 = "v1 original",
                          ms_score_v2 = "v2 TG-corrected",
                          ms_score_v3 = "v3 TG-corr + fixed bounds"))

p_dens <- ggplot(long3, aes(cmdi, colour = version, fill = version)) +
  geom_density(alpha = 0.22) +
  geom_vline(xintercept = CUT["v1"], linetype = 3, colour = "firebrick") +
  geom_vline(xintercept = CUT["v3"], linetype = 2, colour = "darkgreen") +
  labs(x = "CMDI", y = "Density", colour = NULL, fill = NULL) +
  theme(legend.position = "bottom")

p_hist <- ggplot(cmp, aes(ms_score_v3)) +
  geom_histogram(bins = 30, fill = scales::alpha("grey35", 0.4)) +
  geom_vline(xintercept = CUT["v3"], linetype = 2, colour = "darkgreen") +
  labs(x = "CMDI (v3)", y = "Count")

p_dens / p_hist + plot_annotation(tag_levels = "A")
```

The histogram of the adopted construction is unimodal with a slightly longer
lower tail and no pile-up at either boundary. Apparent shoulders in the kernel
density estimate do not persist across bin widths and are attributable to
smoothing rather than structure.

# Zero-collapse audit

```{r}
#| label: zero-audit

zero_count <- function(tag) {
  dd <- cmp |> select(starts_with("d_") & ends_with(tag))
  sum(rowSums(dd <= 0, na.rm = TRUE) > 0)
}

tibble(
  version = c("v1 original", "v2 TG-corrected", "v3 fixed bounds (after floor)",
              "v3 before floor"),
  n_with_zero_component = c(zero_count("_v1"), zero_count("_v2"), zero_count("_v3"),
                            sum(bp_rf_output_reducedIndex_fixed$n_zero_raw > 0,
                                na.rm = TRUE))
) |> kable()
```

| Version | Observations with a zero component |
|---|---:|
| v1 original | 4 |
| v2 TG-corrected | 4 |
| v3 fixed bounds | 0 |
| v3 before the floor | 0 |

The four observations that collapsed to zero under data-driven bounds receive
plausible scores under fixed bounds:

```{r}
#| label: floor-cases
#| eval: true

cmp |>
  filter(ms_score_v1 == 0 | ms_score_v2 == 0 | ms_score_v3 == 0 |
           ID %in% c(207, 132)) |>
  select(ID, Timepoint, ms_score_v1, ms_score_v2, ms_score_v3) |>
  kable(digits = 2)
```

# Rank stability

```{r}
#| label: rank-stability
#| eval: true

pair_stats <- function(a, b, lab) {
  tibble(comparison = lab,
         spearman   = cor(a, b, method = "spearman"),
         kendall    = cor(a, b, method = "kendall"),
         pearson    = cor(a, b, method = "pearson"),
         med_abs_rank_shift = median(abs(rank(b) - rank(a))),
         pct_moving_gt10pct = 100 * mean(abs(rank(b) - rank(a)) > 0.10 * length(a)))
}

rk <- bind_rows(
  pair_stats(cmp$ms_score_v1, cmp$ms_score_v2, "v1 -> v2  (data correction)"),
  pair_stats(cmp$ms_score_v2, cmp$ms_score_v3, "v2 -> v3  (parameterisation)"),
  pair_stats(cmp$ms_score_v1, cmp$ms_score_v3, "v1 -> v3  (total)")
)

rk |> kable(digits = 4)
```

```{r}
#| label: fig-scatter
#| fig-cap: "Participant-level comparison of the three constructions."
#| fig-width: 11
#| fig-height: 5
#| eval: true

theme_set(ggstatsplot::theme_ggstatsplot())

p_v1v2 <- ggplot(cmp, aes(ms_score_v1, ms_score_v2)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_point(alpha = 0.65, size = 1.7, colour = "grey30") +
  labs(x = "CMDI v1 (published)", y = "CMDI v2 (corrected TG)",
       subtitle = "Data correction only")

p_v2v3 <- ggplot(cmp, aes(ms_score_v2, ms_score_v3)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_point(alpha = 0.65, size = 1.7, colour = "steelblue") +
  labs(x = "CMDI v2 (data-driven bounds)", y = "CMDI v3 (fixed bounds)",
       subtitle = "Parameterisation only")

p_v1v3 <- cmp |>
  mutate(changed = str_v1 != str_v3) |>
  ggplot(aes(ms_score_v1, ms_score_v3)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_point(aes(colour = changed), alpha = 0.75, size = 1.8) +
  scale_colour_manual(values = c(`FALSE` = "grey55", `TRUE` = "firebrick"),
                      name = "Changes stratum") +
  labs(x = "CMDI v1 (published)", y = "CMDI v3 (adopted)",
       subtitle = "Total change") +
  theme(legend.position = "bottom")

p_v1v2 | p_v2v3 | p_v1v3
```

Ordering is well preserved across all three comparisons, which is the quantity
that determines whether the refitted model will resemble the published one. The
correction and the reparameterisation each contribute a comparable amount of
movement, and neither dominates.

# Stratum changes

```{r}
#| label: stratum
#| eval: true

stratum_tab <- function(a, b, lab) {
  tibble(comparison = lab,
         n_changed  = sum(a != b),
         pct        = 100 * mean(a != b))
}

bind_rows(
  stratum_tab(cmp$str_v1, cmp$str_v2, "v1 -> v2"),
  stratum_tab(cmp$str_v2, cmp$str_v3, "v2 -> v3"),
  stratum_tab(cmp$str_v1, cmp$str_v3, "v1 -> v3")
) |> kable(digits = 1)

table(from = cmp$str_v1, to = cmp$str_v3) |> kable(caption = "v1 -> v3")
```

Observations closest to the cut-point are those most likely to move again on
refitting, and are listed below for reference.

```{r}
#| label: near-cut
#| eval: true

cmp |>
  mutate(dist = abs(ms_score_v3 - CUT["v3"])) |>
  arrange(dist) |>
  select(ID, Timepoint, ms_score_v1, ms_score_v3, dist) |>
  head(10) |>
  kable(digits = 3)
```

# Component structure and relative influence

## Correlations between components

```{r}
#| label: component-cor
#| eval: true

bp_rf_output_reducedIndex_fixed |>
  select(starts_with("d_")) |>
  cor(method = "spearman") |>
  round(3) |>
  kable(caption = "Spearman correlations between component desirabilities")
```

The components are only weakly associated. The strongest relationship, between
triglycerides and HDL, is consistent with atherogenic dyslipidaemia. Glucose is
close to independent of both lipid components, which is less expected given that
insulin resistance would predict some association along the same axis.

Two implications can be noted. First, it explains why the index is sensitive to how
the components are weighted: with near-orthogonal components there is little
shared variance to stabilise the aggregate. Second, it is worth noting that the
CMDI is a multi-response desirability aggregation rather than a latent-variable
measurement model. The Derringer–Suich framework makes no assumption that the
components load on a common factor; it aggregates desirability, which is a
statement about what is clinically preferable rather than a claim about
dimensionality. This distinguishes the approach from factor-analytic severity
scores, which do assume a single underlying factor.

## Relative influence of each component

Under a geometric mean, each component's contribution to variation in the
composite is governed by the spread of its log-desirability. Since
$\log d_i = s_i \log f_i$, influence is proportional to the scale parameter.

```{r}
#| label: influence

influence <- bp_rf_output_reducedIndex_fixed |>
  select(starts_with("d_")) |>
  summarise(across(everything(), ~ sd(log(pmax(.x, 1e-6))))) |>
  pivot_longer(everything(), names_to = "component", values_to = "log_sd") |>
  mutate(share = log_sd / sum(log_sd))

influence |> kable(digits = 3)
```

| Component | Share of composite variation |
|---|---:|
| BMI | 18.8% |
| TRIG | 25.6% |
| HDL | 46.7% |
| GLU | 8.9% |

The contributions are uneven, with HDL carrying close to half the variation and
glucose under a tenth. This is a consequence of where each anchor falls within
its bounds rather than a deliberate weighting: HDL is the only component whose
anchor sits in the upper part of its range, which produces a convex transform
that expands differences, while the other three have concave transforms that
compress them. For a `d_min` component, widening the upper bound pushes the
anchor fraction toward 1 and inflates the scale parameter sharply, so the
widening adopted for portability necessarily increased HDL's influence.

This is reported as a property of the construction rather than corrected, for
reasons set out in the next section.

# Alternatives considered

Several alternative parameterisations were tested and set aside. Each is recorded
here because the reasons for rejection are part of the justification for the
adopted construction.

| Alternative | Rank correlation with adopted | Reason for rejection |
|---|---:|---|
| Data-driven bounds (`use_data = TRUE`) | — | Scores depend on other participants' data; four observations collapse to zero |
| Mixed anchors (0.5 / 0.6 by inequality type) | 0.989 | Direction defensible, magnitude disproportionate to a single-point distinction |
| Targets solved to equalise contributions | 0.911 | Derived from the sample's own variances, and the IDF thresholds map to desirabilities spanning 0.24–0.76 with no clinical meaning |
| Linear transform (`target` = anchor fraction) | — | Abandons the anchoring entirely; triglycerides then dominate at roughly 47% |
| Narrow symmetric bounds | 0.982 | 43 observations (14.5%) saturate at the HDL ceiling, reintroducing the zero-collapse problem |

Note on two of these:

**Equalising the component contributions** is achievable by solving for the
target that gives each component the required scale. It was rejected on two
grounds. The solution depends on the spread of each component in this particular
sample, so it reintroduces the sample-dependence that motivated the move away
from data-driven bounds, simply relocated from the bounds to the targets. It also
empties the target of clinical meaning: a participant at the glucose threshold
would be scored as less metabolically undesirable than one at the HDL threshold,
by a factor of three, with no rationale. There is also a tension with the
argument made in the manuscript's introduction, which criticises binary
definitions for treating criteria as equivalent despite heterogeneous risk
implications; equal weighting by construction does the same thing.

**Narrow symmetric bounds** are attractive because bounds symmetric about the
anchor give a scale of exactly 1, producing a linear transform with the threshold
at the midpoint. The rank correlation with the adopted construction was 0.982,
which initially looked acceptable. The saturation audit showed why it was not:
43 observations sat at or above an HDL ceiling of 70 mg/dL, a value that is
common rather than extreme. The high rank correlation persisted because the
clamping is directionally correct — high HDL is protective and those
participants belong at the low-risk end — but their HDL information is lost.
This is a useful reminder that rank correlation alone does not detect degraded
measurement.

# Summary and remaining checks

The adopted construction uses corrected triglycerides, fixed clinical bounds, a
uniform anchor of 0.5, and a small floor applied before aggregation. On the
evidence assembled here:

- No component can reach zero, so the geometric-mean collapse that produced
  implausible scores under data-driven bounds cannot occur.
- Saturation is limited to one observation at the triglyceride ceiling, in a
  range where clinical management does not differentiate.
- Participant ordering is well preserved relative to the published index, so the
  refitted model is expected to resemble the published one in structure, with
  movement concentrated among predictors near the selection threshold.
- The index no longer depends on which participants happen to be in the sample,
  which is a prerequisite for the external validation described in the
  manuscript.
- Component contributions are uneven and this is disclosed rather than adjusted,
  since every attempt to balance them cost more in interpretability or
  portability than it gained.

A check before refitting:

The near-zero correlation between glucose and the lipid components is not what
insulin-resistance pathophysiology would predict, particularly since the
triglyceride–HDL association is present. Given that an error has already been
found in this panel, it is worth confirming that fasting status was standardised
and recorded, and that glucose was assayed from the same draw as the lipids.
Inconsistent fasting would decorrelate glucose while leaving the lipid
relationships intact, which matches the observed pattern.

```{r}
#| label: session

sessionInfo()
```
